### 2020-06-03 writeup/setup.tex
revisions

Equation \eqref{eq_te_model_shift} corresponds to the bias of $\hat{\beta}_t^{\MTL}$, which captures how similar $\beta_1$ and $\beta_2$ are.
Hence, the bias of $\hat{\beta}_t^{\MTL}$ introduces a negative effect from adding the source labels.
Equation \eqref{eq_te_var} corresponds to the variance of $\hat{\beta}_t^{\MTL}$, which is always smaller than the variance of $\hat{\beta}_t^{\STL}$.
%This part introduces a positive variance reduction effect from adding the source labels.
Hence, whether $\te(\hat{\beta}_t^{\MTL}) < \te(\hat{\beta}_t^{\STL})$ is determined precisely by the tradeoff between the negative effect of the bias term and the positive effect of the variance term!
%(i) the negative effect from model shift bias.
%(ii) the positive effect from variance reduction;

@@

%\textbf{Multi-task learning.}
We provide a sharp analysis of the bias-variance tradeoff two tasks with general covariance matrices.
We state our result for two tasks as follows.
%The formal statement is stated in Theorem \ref{thm_many_tasks} and its proof can be found in Appendix \ref{app_proof_many_tasks}.
%The technical crux of our approach is to derive the asymptotic limit of $\te(\hat{\beta}_t^{\MTL})$ in the high-dimensional setting, when $p$ approaches infinity.
%We derive a precise limit of $\bigtr{(X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}\Sigma_2}$, which is a deterministic function that only depends on $\Sigma_1, \Sigma_2$ and $n_1/p, n_2/p$ (see Lemma \ref{lem_cov_shift} in Appendix \ref{app_proof_main} for the result).
%Based on the result, we show how to determine positive versus negative transfer as follows.
%, where $\lambda_{\min}(M)$ is the smallest singular value of $M_1$
\begin{corollary}\label{thm_main_informal}
	For the setting of two tasks, let $M=\Sigma_1^{1/2}\Sigma_2^{-1/2}$, $\delta > 0$ be a desired error margin and $\rho_1 \gtrsim \frac{1}{\delta^2}\cdot \lambda_{\min}(M)^{-4} \norm{\Sigma_1} \max(\norm{\beta_1}^2, \norm{\beta_2}^2)$.
	Let $\rho_2 > 1$ be a fixed value.
 	There exists two deterministic functions $\Delta_{\bias}$ and $\Delta_{\vari}$ that only depend on $\set{\hat{v}, \Sigma_1, \Sigma_2, \rho_1, \rho_2, \beta_1, \beta_2}$ such that
	\squishlist
		\item If $\Delta_{\bias} - \Delta_{\vari} < -\delta$, then w.h.p. over the randomness of $X_1, X_2$, we have $\te(\hat{\beta}_t^{\MTL}) < \te(\hat{\beta}_t^{\STL})$.
		\item If $\Delta_{\bias} - \Delta_{\vari} > \delta$, then w.h.p. over the randomness of $X_1, X_2$, we have $\te(\hat{\beta}_t^{\MTL}) > \te(\hat{\beta}_t^{\STL})$.
	\squishend
\end{corollary}

Corollary \ref{thm_main_informal} applies to settings where large amounts of source task data is available but the target data size is small.
For such settings, we obtain a sharp transition from positive transfer to negative transfer determined by $\Delta_{\bias} - \Delta_{\vari}$.
%determined by the covariate shift matrix and the model shift.
%The bounds get tighter and tighter as $\rho_1$ increases.
While the general form of the threshold can be complex (as is preivous generalization bounds for MTL), they admit interpretable forms for simplified settings.
This will be the focus of Section \ref{sec_insight}.

\textbf{Proof Overview.} We first describe the proof of Theorem \ref{lem_cov_shift_informal}.
We use the Stieltjes transform method (or the resolvent method) in random matrix theory \cite{bai2009spectral,tao2012topics,erdos2017dynamical}. Roughly speaking, we study the resolvent $R(z):=[\Sigma_2^{-1/2}( X_1^{\top}X_1 + X_2^{\top}X_2)\Sigma_2^{-1/2}-z]^{-1}$ for $z\in \C$ around $z=0$.
Using the methods in \cite{Anisotropic,yang2019spiked}, we find the asymptotic limit, say $R_\infty(z)$, of $R(z)$ for any $z$ as $p\to \infty$ with an almost optimal convergence rate. In particular, when $z=0$, $\tr[R_\infty(0)]$ gives the right hand side of \eqref{eq_introX1X2}, which concludes Theorem \ref{lem_cov_shift_informal}. The details can be found in Appendix \ref{sec_maintools} and \ref{sec_Gauss}.

Using Theorem \ref{lem_cov_shift_informal} over \eqref{eq_te_var}, we can calculate the amount of reduced variance compared to STL.
The amount of reduced variance is given by $\Delta_{\vari}$.
For the bias term of equation \eqref{eq_te_model_shift}, we apply a Gaussian concentration bound on $X_1^{\top}X_1$, whose expectation is $n_1^2\Sigma_1$.
This results in the error term $\delta$, which scales as $(1 \pm 1/\sqrt{\rho_1})^4$.
Then, we applying a similar identity to Theorem \ref{lem_cov_shift_informal} for bounding the bias term, noting that the derivative of $R(z)$ with respect to $z$ can be approximated by $R_\infty'(z)$.
This leads to a negative effect given by $\Delta_{\bias}$. %, which will be used to estimate the first term on the right hand side of \eqref{eq_te_model_shift}.
%During this process, we will get the $\Delta_{\bias}$ term up to an error $\delta$ depending on $\rho_1$.
The details are presented in Appendix \ref{app_proof_main}.

%The proof of Theorem \ref{thm_main_informal} is presented in Appendix \ref{app_proof_main}.
Next, we describe our result for more than two tasks with same features, i.e. $X_i = X$ for any $i$.
This setting is prevalent in applications of multi-task learning to image classification, where there are multiple prediction labels/tasks for every image \cite{chexnet17,EA20}.
\begin{theorem}\label{thm_many_tasks}
%Suppose $X=Z\Sigma^{1/2}\in \R^{n\times p}$ satisfy Assumption \ref{assm_secA1} with $\rho:=n/p>1$ being some fixed constant. Consider data models  $Y_i = X\beta_i + \varepsilon_i$, $i=1,2,\cdots, t$, where $\e_i\in \R^{n}$ are random vectors with i.i.d. entries with mean zero, variance $\sigma^2$ and all moments as in \eqref{assmAhigh}. Moreover, assume that $X$, $\beta_i$ and $\e_i$ are all independent of each other.
	%Let $n = c \cdot p$.
	%Let $X\in\real^{n\times p}$ and $Y_i = X\beta_i + \varepsilon_i$, for $i = 1,\dots,k$.
	Consider $t$ data models $Y_i = X\beta_i + \varepsilon_i$, $i=1,2,\cdots, t$, that satisfy Assumption \ref{assm_secA2}.
	Let $U_r U_r^{\top}$ denote the best rank-$r$ subspace approximation of $(B^{\star})^\top\Sigma B^{\star}$, where $B^\star := [{\beta}_1,{\beta}_2,\dots,{\beta}_{t}]$ and $U_r\in\real^{t\times r}$. Suppose $(B^{\star})^\top\Sigma B^{\star}$ is of full rank in the sense that $\lambda_{\min}((B^{\star})^\top\Sigma B^{\star})\gtrsim \sigma^2$. Let $U_r(i)$ denote the $i$-th row vector of $U_r$.
	We have the following
	\squishlist
		\item We have $\te(\hat{\beta}_t^{\MTL}) < \te(\hat{\beta}_t^{\STL})$ with high probability if
		$$\left(1 - \norm{U_r(t)}^2 \right)\frac{\sigma^2}{\rho - 1} > \norm{\Sigma (B^{\star} U_r U_r(t) - \beta_t)}^2 +  \oo \left( \|B^\star\|^2 + \sigma^2\right).$$
		\item We have $\te(\hat{\beta}_t^{\MTL}) > \te(\hat{\beta}_t^{\STL})$ with high probability if
		$$\left(1 - \norm{U_r(t)}^2\right)\frac{\sigma^2}{\rho - 1} < \norm{\Sigma(B^{\star} U_r U_r(t) - \beta_t)}^2 -  \oo \left( \|B^\star\|^2 + \sigma^2\right).$$
	\squishend
\end{theorem}
%A similar result for the second setting can be found in Appendix \ref{app_proof_many_tasks}.

\textbf{Remark.} Theorem \ref{lem_cov_shift_informal} extends a well-known result for the single-task setting when $X_1, \rho_1, a_1$ are all equal to zero \cite{S07}.
The concentration error of our result that is order $O(p^{-1/2+o(1)})$ is nearly optimal.

>>>>

The term on the right-hand side of \eqref{eq_te_model_shift} corresponds to the bias of $\hat{\beta}_t^{\MTL}$, which captures how similar $\beta_1$ and $\beta_2$ are.
Hence, the bias of $\hat{\beta}_t^{\MTL}$ introduces a negative effect from adding the source labels.
Equation \eqref{eq_te_var} corresponds to the variance of $\hat{\beta}_t^{\MTL}$, which is always smaller than the variance of $\hat{\beta}_t^{\STL}$.
%This part introduces a positive variance reduction effect from adding the source labels.
Hence, whether $\te(\hat{\beta}_t^{\MTL}) < \te(\hat{\beta}_t^{\STL})$ is determined precisely by the tradeoff between the negative effect of the bias term and the positive effect of the variance term!
%(i) the negative effect from model shift bias.
%(ii) the positive effect from variance reduction;

@@

%\textbf{Multi-task learning.}
We provide a sharp analysis of the bias-variance tradeoff for two tasks with general covariance matrices.
We state our result as follows.
%The formal statement is stated in Theorem \ref{thm_many_tasks} and its proof can be found in Appendix \ref{app_proof_many_tasks}.
%The technical crux of our approach is to derive the asymptotic limit of $\te(\hat{\beta}_t^{\MTL})$ in the high-dimensional setting, when $p$ approaches infinity.
%We derive a precise limit of $\bigtr{(X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}\Sigma_2}$, which is a deterministic function that only depends on $\Sigma_1, \Sigma_2$ and $n_1/p, n_2/p$ (see Lemma \ref{lem_cov_shift} in Appendix \ref{app_proof_main} for the result).
%Based on the result, we show how to determine positive versus negative transfer as follows.
%, where $\lambda_{\min}(M)$ is the smallest singular value of $M_1$
\begin{corollary}\label{thm_main_informal}
	For the setting of two tasks, let $M=\Sigma_1^{1/2}\Sigma_2^{-1/2}$, $\delta > 0$ be a desired error margin, $\rho_1 \gtrsim \delta^{-2}\cdot \lambda_{\min}(M)^{-4} \norm{\Sigma_1} \max(\norm{\beta_1}^2, \norm{\beta_2}^2)$, and $\rho_2 > 1$ be a fixed value.
 	There exists two deterministic functions $\Delta_{\bias}$ and $\Delta_{\vari}$ that only depend on $\set{\hat{v}, \Sigma_1, \Sigma_2, \rho_1, \rho_2, \beta_1, \beta_2}$ such that
	\squishlist
		\item If $\Delta_{\bias} - \Delta_{\vari} < -\delta$, then w.h.p. over the randomness of $X_1, X_2$, we have $\te(\hat{\beta}_t^{\MTL}) < \te(\hat{\beta}_t^{\STL})$.
		\item If $\Delta_{\bias} - \Delta_{\vari} > \delta$, then w.h.p. over the randomness of $X_1, X_2$, we have $\te(\hat{\beta}_t^{\MTL}) > \te(\hat{\beta}_t^{\STL})$.
	\squishend
\end{corollary}

Corollary \ref{thm_main_informal} applies to settings where large amounts of source task data is available but the target data size is small.
For such settings, we obtain a sharp transition from positive transfer to negative transfer determined by $\Delta_{\bias} - \Delta_{\vari}$.
%determined by the covariate shift matrix and the model shift.
%The bounds get tighter and tighter as $\rho_1$ increases.
While the general form of the threshold can be complicated (as are the previous generalization bounds for MTL), they admit interpretable forms for simplified settings.
This will be the focus of Section \ref{sec_insight}.

\textbf{Proof Overview.} We first describe the proof of Theorem \ref{lem_cov_shift_informal}.
We use the Stieltjes transform method (or the resolvent method) in random matrix theory \cite{bai2009spectral,tao2012topics,erdos2017dynamical}. Roughly speaking, we study the resolvent $R(z):=[\Sigma_2^{-1/2}( X_1^{\top}X_1 + X_2^{\top}X_2)\Sigma_2^{-1/2}-z]^{-1}$ for $z\in \C$ around $z=0$.
Using the methods in \cite{Anisotropic,yang2019spiked}, we find the asymptotic limit, say $R_\infty(z)$, of $R(z)$ for any $z$ as $p\to \infty$ with an almost optimal convergence rate. In particular, when $z=0$, $\tr[R_\infty(0)]$ gives the expression inTheorem \ref{lem_cov_shift_informal}. The details can be found in Appendix \ref{sec_maintools} and \ref{sec_Gauss}.

Applying Theorem \ref{lem_cov_shift_informal} to \eqref{eq_te_var}, we can calculate the amount of reduced variance compared to STL, which is given asymptotically by $\Delta_{\vari}$.
For the bias term in equation \eqref{eq_te_model_shift}, we apply the approximate isometry property to $X_1^{\top}X_1$, which is close to $n_1^2\Sigma_1$. This results in the error term $\delta$, which scales as $(1 + 1/\sqrt{\rho_1})^4-1$.
Then, we apply a similar identity to Theorem \ref{lem_cov_shift_informal} for bounding the bias term, noting that the derivative of $R(z)$ with respect to $z$ can be approximated by $R_\infty'(z)$.
This estimates the negative effect given by $\Delta_{\bias}$. %, which will be used to estimate the first term on the right hand side of \eqref{eq_te_model_shift}.
%During this process, we will get the $\Delta_{\bias}$ term up to an error $\delta$ depending on $\rho_1$.
The details are presented in Appendix \ref{app_proof_main}.

%The proof of Theorem \ref{thm_main_informal} is presented in Appendix \ref{app_proof_main}.
Next, we describe our result for more than two tasks with same features, i.e. $X_i = X$ for any $i$.
This setting is prevalent in applications of multi-task learning to image classification, where there are multiple prediction labels/tasks for every image \cite{chexnet17,EA20}.
\begin{theorem}\label{thm_many_tasks}
%Suppose $X=Z\Sigma^{1/2}\in \R^{n\times p}$ satisfy Assumption \ref{assm_secA1} with $\rho:=n/p>1$ being some fixed constant. Consider data models  $Y_i = X\beta_i + \varepsilon_i$, $i=1,2,\cdots, t$, where $\e_i\in \R^{n}$ are random vectors with i.i.d. entries with mean zero, variance $\sigma^2$ and all moments as in \eqref{assmAhigh}. Moreover, assume that $X$, $\beta_i$ and $\e_i$ are all independent of each other.
	%Let $n = c \cdot p$.
	%Let $X\in\real^{n\times p}$ and $Y_i = X\beta_i + \varepsilon_i$, for $i = 1,\dots,k$.
	Consider $t$ data models $Y_i = X\beta_i + \varepsilon_i$, $i=1,2,\cdots, t$, where $X$ has covariance matrix $\Sigma$, and the entries of $\e_i$ are i.i.d. with mean zero and variance $\sigma^2$.  
	%that satisfy Assumption \ref{assm_secA2} in the appendix.
	Let $U_r U_r^{\top}$ denote the best rank-$r$ subspace approximation of $(B^{\star})^\top\Sigma B^{\star}$, where $B^\star := [{\beta}_1,{\beta}_2,\dots,{\beta}_{t}]$ and $U_r\in\real^{t\times r}$. Suppose $(B^{\star})^\top\Sigma B^{\star}$ is of full rank in the sense that $\lambda_{\min}((B^{\star})^\top\Sigma B^{\star})\gtrsim \sigma^2$. Let $U_r(i)$ denote the $i$-th row vector of $U_r$.
	We have the following
	\squishlist
		\item $\te(\hat{\beta}_t^{\MTL}) < \te(\hat{\beta}_t^{\STL})$ with high probability if
		$$\left(1 - \norm{U_r(t)}^2 \right)\frac{\sigma^2}{\rho - 1} > \norm{\Sigma (B^{\star} U_r U_r(t) - \beta_t)}^2 +  \oo \left( \|B^\star\|^2 + \sigma^2\right).$$
		\item $\te(\hat{\beta}_t^{\MTL}) > \te(\hat{\beta}_t^{\STL})$ with high probability if
		$$\left(1 - \norm{U_r(t)}^2\right)\frac{\sigma^2}{\rho - 1} < \norm{\Sigma(B^{\star} U_r U_r(t) - \beta_t)}^2 -  \oo \left( \|B^\star\|^2 + \sigma^2\right).$$
	\squishend
\end{theorem}
%A similar result for the second setting can be found in Appendix \ref{app_proof_many_tasks}.

\textbf{Remark.} Theorem \ref{lem_cov_shift_informal} extends a well-known result for the single-task setting when $X_1, \rho_1, a_1$ are all equal to zero \cite{S07}.
The concentration error $\OO(p^{-1/2+o(1)})$ of our result is nearly optimal.
